Calibrating seed-based heuristics to map short reads with Sesame

### 2020-03-26 manuscript.tex
State of the manuscript on resubmission

\title{Calibrating seed-based mapping heuristics with Sesame}

@@

\begin{definition}
\label{def_dup}
The target is the DNA fragment that was actually sequenced. Duplicates are
sequences of the genome that share homology with the target (in genetics
they are often referred to as paralogs).
\end{definition}

The difficulty of the true location problem is due to sequencing errors.
Occasionally, the sequence of a DNA fragment can be closer to one of the
duplicates than to the target. So the \emph{true} location of a DNA
fragment is not always the \emph{best} (as measured by the identity
between the sequence and the candidate location). This naturally leads to
ask how we can identify the \emph{best} location of the fragment in
the genome.

@@

Exact alignment algorithms were designed to solve the best location
problem~\cite{pmid7265238, pmid5420325}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.}, algorithms that run faster but may
return an incorrect result~\cite{Waterman1984}.

The most popular heuristic for mapping DNA sequences is a filtration
method called ``seed and extend''. The principle is to first identify
seeds, defined as short regions of high similarity between the read and
the genome, and then use an exact alignment algorithm at the seeded
locations to evaluate the candidates and identify the optimum. This
strategy was first proposed in FASTA~\cite{pmid2983426} and
BLAST~\cite{pmid2231712} to to search for homology in sequence databases
of proteins and DNA.

@@

\begin{definition}
The candidate set obtained at the seeding step is said to be
\begin{enumerate}[i)]
\item ``on target'' if it contains the target,
\item ``off target'' if it contains a duplicate but not the target, 
\item ``null'' if it contains neither.
\end{enumerate}
\end{definition}

With our assumptions, a read is mapped to the wrong location if and only
if the candidate set is off target. Indeed, if the candidate set is null,
the read is not mapped, and if the candidate set is on target, the correct
location is discovered at the alignment step. The equivalence is granted
by the assumption that the true location is also the best, reducing the
mapping problem to a seeding problem. That being said, the true location
is not always the best in practical applications. We will show in
section~\ref{sec:truevsbest} that off-target seeding is responsible for
most of the mapping errors even without the assumption above, but for now
we maintain the strict equivalence between mapping errors and off-target
seeding.

Focusing on popular heuristics to map short reads, our aim is to develop a
method to estimate the probability that the candidate set is off target.
Previous work pioneered a method to compute seeding probabilities but it
did not distinguish off-target from null
seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
not provide a way to estimate the frequency of mapping errors. Other
authors investigated the reliability of mapping
algorithms~\cite{pmid23872968}, but they focused on the probabilities of
random hits, recognizing that addressing the problem of incorrect mapping
requires taking into consideration the repeat structure of the genome.

The rest of the article is organized as follows: section~\ref{sec_seeds}
presents common seeding strategies used in bioinformatics,
section~\ref{sec:symbolic} presents the basic concepts of analytic
combinatorics that will be required to compute seeding probabilities,
sections~\ref{sec:offexact} to \ref{sec_offMEM} treat the cases of exact
seeds, skip seeds and MEM seeds, three common seeding heuristic used for
mapping, section~\ref{sec_practical} returns to the mapping problem and
revisits the assumptions of the model, section~\ref{sec:sesame} presents
Sesame, a C library implementing the main results of the theory, and
finally section~\ref{sec_disc} provides some perspectives on the present
work.

@@

\begin{definition}
A seed is a subsequence of the read that has size at least $\gamma$
(defined by the context of the problem) and that has at least one perfect
match in the reference genome.
\end{definition}

@@

The success of the mainstream mappers BWA-MEM~\cite{li2013aligning} and
Bowtie2~\cite{pmid22388286} is due in part to the FM-index, which only
supports contiguous seeds. Some workarounds are
available~\cite{horton2008dislex, gagie2017compressed} but they increase
the memory footprint, explaining that short reads are typically mapped
using contiguous seeds. More generally, computing the sensitivity of
spaced seeds is challenging~\cite{pmid16819802, li2006superiority,
martin2017faster}. It is theoreticlally possible to do this using the
tools introduced below, but the complexity becomes prohibitive for seed
models with several don't-care positions. For these reasons we will not
treat the case of spaced seeds and we will give more detail in
section~\ref{sec_disc} bout the kind of difficulties they represent for
our approach.

@@

Exact seeds are seeds of fixed size $\gamma$. In other words, when using
exact seeds, the candidate set consists of all the genomic locations for
which there is a perfect match of size $\gamma$ in the read. This seeding
heuristic was used in the first version of BLASTN~\cite{pmid2231712}, but
it has become unpopular for producing many redundant hits within the same
regions of the read and of the genome.

Figure~\ref{fig:exact_seed_example} shows the exact seeds from an example
a read with three miscalled nucleotides. The sequenced DNA fragment has
three duplicates so the seeds can match four possible locations.

@@

Before going further, we also need to move a practical consideration out
of the way. Seeds can match any sequence of the genome, not just the
target or the duplicates. However, we will ignore matches in the rest of
the genome because such random matches are unlikely to cause a mapping
error when seeding is off target, contrary to matches in duplicates.
Neglecting those will greatly simplify the exposition of the theory
without loss of generality. We will explain in
section~\ref{sec:random_seeds} how to deal with this practical case and
and how to identify those matches as off target. Until then, we will
consider that the target and the duplicates are the only sequences in the
reference genome.

@@

To give an example, assume that a particular symbol, say $\Downarrow$, has
a probability of occurence equal to $p$. The weighted generating function
words containing only this symbol is $pz$. The weight of the word is
its probability (here equal to $p$) and the size is its length in number
of symbols (here 1).

In this document we will focus on the weighted generating function $A(z)$
of the set $\mathcal{A}$ of reads that do not have on-target seeds
(\textit{i.e.} reads for which seeding is either null or off target). The
weight of a read is its probability of occurrence and the size $k$ is its
number of nucleotides. The coefficient $a_k$ is thus the proportion of
reads of size $k$ that do not have an on-target seed, which is the
quantity of interest.

@@

The analytic combinatorics framework relies on a strategy referred to as
the \emph{symbolic method}~\cite{sedgewick2013introduction}. The idea is
to combine simple objects into more complex objects. Each combinatorial
operation on the objects corresponds to a mathematical operation on their
weighted generating functions. One can thus obtain the weighted generating
function of complex objects, whose coefficients $a_k$ are the
probabilities of interest~\cite{regnier2000unified, nicodeme2002motif,
sedgewick2013introduction}.

@@

We define the mismatch alphabet $\mathcal{A}_0 = \{\square, |,
\Downarrow\}$, where $\square$ represents a correct nucleotide,
$\Downarrow$ represents an erroneous nucleotide and $|$ is special symbol
appended after the last nucleotide to mark the end of the read. It is not
associated to a nucleotide and therefore has size 0.

@@

The probability of occurrence of the symbol $\Downarrow$ is $p$ (the error
rate of the sequencer) so the probability of occurrence of the $\square$
symbol is $1-p = q$. Both symbols have size 1, so their respective
weighted generating function are $pz$ and $qz$. Using the rule for
concatenation, we see that a segment of $i$ symbols $\square$ followed by
a terminator has weighted generating function $(qz)^ipz$. The final symbol
$|$ has size 0, so a tail segment of $i$ symbols $\square$ followed by the
symbol $|$ has weighted generating function $(qz)^i$.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_skip.pdf}
\caption{\textbf{The skip encoding.}
The read of Figure~\ref{fig:simple} is represented in the skip-3 alphabet.
The symbols $\Downarrow_j (j = 1,2,3)$ represent mismatches against the
target (they are erroneous nucleotides) and the $\square$ symbol
represents a match (it is a correct nucleotide). The vertical bars
indicate non-skipped positions (the potential start of a seed). The number
$j$ in $\Downarrow_j$ indicates the number of nucleotidesu until the next
non-skipped position. For $j=0$ the next nucleotide is not skipped.
Other features are as in Figure~\ref{fig:simple}.}
\label{fig:skip}
\end{figure}

@@

By construction, the Taylor expansion of the top right term in the matrix
$M_n(z)\cdot(I-M_n(z))^{-1}$ contains the the probabilities of interest.
More specifically, if the Taylor series of this term is $a_0 + a_1 z + a_2
z^2 + \ldots$, then $a_k$ is the probability that a read of size $k$
contains no skip-$n$ seed of minimum size $\gamma$.

@@

We now turn our attention to the problem of computing the probability that
the seeding process is off target when using exact seeds---recall from
section~\ref{sec:twotypes} that off-target seeding means that the
candidate set contains a duplicate but not the target.

If there is no duplicate (\textit{i.e.} $N = 0$), seeding cannot be off
target, it can only be on target or null. So from here we assume that the
target has $N \geq 1$ duplicates. Let $S_0$ denote the event that there is
an on-target seed and let $S_j$ denote the event that there is a seed for
the $j$-th duplicate. We are thus interested in computing
$P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big)$, where
$\overline{S}_j$ denotes the complement of the event $S_j$. First observe
that
\begin{equation}
\label{eq:pcomp}
P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big) =
P(\overline{S}_0) - P\big(\overline{S}_0 \cap \overline{S}_1 \cap \ldots
\cap \overline{S}_N\big).
\end{equation}

Since the duplicates are assumed to evolve independently of each other and
through the same mutagenesis process, the events $S_j$ ($1 \leq j \leq N$)
are independent and identically distributed conditionally on
$\overline{S}_0$. We can thus write
\begin{eqnarray}
\notag
P(\overline{S}_0 \cap \ldots \cap \overline{S}_N) &=&
  P(\overline{S}_0) \cdot P(\overline{S}_1 \cap \ldots \cap
  \overline{S}_N | \overline{S}_0) \\
\label{eq:Pnull}
  &=& P(\overline{S}_0) \cdot P(\overline{S}_1 | \overline{S}_0)^N =
  P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0 \cap
\overline{S}_1)}{P(\overline{S}_0)} \right)^N.
\end{eqnarray}
Combinging the two equations above, we obtain
\begin{equation}
\label{eq:Poff}
P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big) =
P(\overline{S}_0) - P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0
\cap \overline{S}_1)}{P(\overline{S}_0)} \right)^N.
\end{equation}

Hence, the probability that seeding is off target is a function of just
two quantitites: $P(\overline{S}_0)$ and $P(\overline{S}_0 \cap
\overline{S}_1)$. The first is the probability that the read has no seed
for the target, which we have already computed in
section~\ref{sec:example_exact} using recursive expression
(\ref{eq:recur}). We now need to find a way to compute $P(\overline{S}_0
\cap \overline{S}_1)$.

\begin{remark}
Observe that expression (\ref{eq:Pnull}) is the probability of null
seeding (the read contains no seed for the target or any of its
duplicates). Since it is also a function of just $P(\overline{S}_0)$ and
$P(\overline{S}_0 \cap \overline{S}_1)$, it can be computed at no
additional cost. This probablity is less relevant than the probabilities
that the seeding process is on target or off target, but at times, it may
be useful to know the probability that a read is not mappable, especially
when reads are relatively short.
\end{remark}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.65]{curves_exact.pdf}
\caption{\textbf{Off-target seeding probabilities (exact seeds).} The
surfaces show the probability that seeding is off target as a function of
the divergence rate $\mu$ and of the number $N$ of duplicates. The results
are shown for exact seeds of size $\gamma=19$ and for reads of size $k=50$
nucleotides sequenced with an error rate $p=0.01$. The divergence rate
$\mu$ is defined as the probability that a given duplicate differs from
the target at any given position. Note the factor 1000 in the scale of the
z-axis.}
\label{fig:curves_exact}
\end{figure}

Figure~\ref{fig:curves_exact} shows the result for a number of duplicates
$N$ from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The first
observation is that the probability that seeding is off target increases
with $N$. This is also clear from expression (\ref{eq:Poff}). This can
also be understood intuitively because the probability of not seeding the
target is fixed and the probability of having an empty candidate set
decreases as $N$ increases. As a result the probability that the candidate
set contains only invalid candidates increases with $N$.

@@

To compute the probability that the seeding process is off target when
using skip seeds, we observe that the logic of section~\ref{sec:offexact}
can be transposed with few modifications. In particular, the probability
can be computed through expression (\ref{eq:Poff}), where $S_0$ is the
event that the read has a \emph{skip} seed for the target (instead of an
exact seed) and $S_1$ is the event that the read has a \emph{skip} seed
for the first duplicate (instead of an exact seed).

@@

For the last two types of terminators, we must pay attention to the fact
that in general, a mismatch against $+$ that follows a mismatch against
$-$ can be represented by either a $\Downarrow_j$ symbol ($0 \leq j \leq
n$) or a $\downarrow_{/j}^+$ symbol ($j \geq 1$). The terminator is
$\Downarrow_j$ if the two mismatches are within the same interval between
non-skipped positions because both sequences lovally have match length 0.
If the mismatch against $+$ is in another interval, the terminator is
$\downarrow_{/j}^+$ because in that case the $-$ sequence has a positive
match length. For now, bear in mind that a mismatch against the $+$
sequence only can produce a $\Downarrow_j$ terminator ($0 \leq j \leq n$),
this will be important later.

@@

We now have all the elements to specify the transfer matrix of reads with
no skip-$n$ seed for either sequence. Recall that $n$ is the number of
skipped nucleotides, $\gamma$ is the minimum seed length, $a$ is the
probabiliy of a double match, $b$ is the probability of a mismatch against
the $-$ sequence only, $c$ is the probability of a mismatch against the
$+$ sequence only and $d$ is the probability of a double mismatch. For
notational convenience we define
\begin{eqnarray}
\label{eq_N}
N_i(z) &=& 1+z+\ldots+z^i, \\
W_j(z) &=& (az)^x \big( 1 + (az)^{n+1} + \ldots +
  (az)^{(n+1)m} \big) dz, \\
  \notag
  &\quad& \text{where } x = -j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma-1-x}{n+1} \right\rfloor, \\
U_{i,j}(z) &=& (az)^x
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
\begin{cases}
  bz\cdot(az)^x \text{ if } j < y, \\
  0 \text{ otherwise}
\end{cases} \\
  \notag
  &\quad& \text{where } x = -i-j-1 \modulo{n+1},
  y = -i \modulo{n+1}, \\
  \notag
  &\quad& \text{and } m = \left\lfloor \frac{\gamma-1-i-x}{n+1}
    \right\rfloor, \\
V_{i,j}(z) &=& (az)^x 
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
\begin{cases}
  cz\cdot(az)^x \text{ if } j \leq x, \\
  0 \text{ otherwise}
\end{cases} \\
  \notag
  &\quad& \text{ where } x = -i-j-1 \modulo{n+1},
  m = \left\lfloor \frac{\gamma-1-i-x}{n+1} \right\rfloor.
\end{eqnarray}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.47]{curves_skip.pdf}
\caption{\textbf{Off-target seeding probabilities (skip seeds).}
The surfaces show the probability that seeding is off target for skip-5
and skip-9 seeds of size $\gamma=19$. Read size and error rate are the
same as in Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and
$p=0.01$. The divergence rate $\mu$, defined as the probability that a
given duplicate differs from the target at any given position. Note the
factor 1000 in the scale of the z-axis.}
\label{fig:curves_skip}
\end{figure}

Figure~\ref{fig:curves_skip} shows the result for a number of duplicates $N$
from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The surfaces
have the same general aspect as in Figure~\ref{fig:curves_exact}. The
probability that seeding is off target increases with $N$ and there is
again a  worst value of $\mu$, because the maximum of $P(\overline{S}_0
\cap \overline{S}_1)$ minimizes expression (\ref{eq:Poff}) for every value
of $N$. However, those values of $\mu$ are not constant (they are
approximately 0.065 and 0.060 for skip-5 and skip-9 seeds, respectively).

Importantly, Figure~\ref{fig:curves_skip} reveals that skipping 5
nucleotides increases the chances that seeding is off target by a factor
approximately 1.5, and skipping 9 nucleotides by a factor approximately
1.8 (compare with Figure~\ref{fig:curves_exact}). Skipping nucleotides
seems to have little effect, but this is not a general conclusion. Indeed,
skipping nucleotides decreases the probability of on-target seeding
(skipping positions implies fewer on-target seeds) but increases the
probability of null seeding (skipping positions implies fewer seeds
overall), so the effects must be evaluated on a case-by-case basis.
Skipping nucleotides can even decrease the probability that seeding is
off target. Concretely, for exact seeds of size $\gamma=19$ in reads of
size $k=50$ with $N=10$ duplicates at divergence $\mu=0.1$ with a
sequencing error $p=0.1$, the probability of off-target seeding is
approximately 0.178 with exact seeds and 0.035 with skip-9 seeds, showing
skipping nucleotides can have different effects.

>>>>

\title{Calibrating seed-based heuristics to map short reads with Sesame}

@@

\begin{definition}
\label{def_dup}
The target is the DNA fragment that was actually sequenced. Duplicates are
sequences of the genome that share homology with the target (in genetics
they are often referred to as paralogs). In this article we will focus on
short reads from complex eukaryotic genomes, so for concreteness the
reader can assume that fragments are 30-300 bp long and that duplicates
have above 75\% identity with the target.
\end{definition}

The difficulty of the true location problem is due to sequencing errors.
Occasionally, the sequence of a DNA fragment can be closer to one of the
duplicates than to the target. So the \emph{true} location of a DNA
fragment is not always the \emph{best} (as measured by the identity
between the sequence and the candidate location). This naturally leads to
asking how we can identify the \emph{best} location of the fragment in
the genome.

@@

Exact alignment algorithms were designed to solve the best location
problem~\cite{pmid5420325, pmid7265238}, but they are too slow to process
the large amount of data generated by modern sequencers. Instead, one uses
heuristic methods, \textit{i.e.}, algorithms that run faster but may
return an incorrect result~\cite{Waterman1984}.

The most popular heuristic for mapping DNA sequences is a filtration
method called ``seed-and-extend''. The principle is to first identify
seeds, defined as short regions of high similarity between the read and
the genome, and then use an exact alignment algorithm at the seeded
locations to evaluate the candidates and identify the optimum. Exact
alignment algorithms, such as the Needleman-Wunsch~\cite{pmid5420325} or
the Smith-Waterman~\cite{pmid7265238} algorithms return the distance or
the similarity between two sequences according to some quantitative
criterion. They are exact in the sense that they always return the correct
answer, unlike heuristic algorithms. The seed-and-extend strategy was
first proposed in FASTA~\cite{pmid2983426} and BLAST~\cite{pmid2231712} to
search for homology in sequence databases of proteins and DNA.

@@

\begin{definition}
The output of the seeding step is the candidate set. The candidate set is
the list of genomic locations where the read can be potentially mapped.
The read is always mapped to one element of the candidate set. The seeding
step is said to be
\begin{enumerate}[i)]
\item ``on target'' if the candidate set contains the target,
\item ``off-target'' if the candidate set contains a duplicate but not the target, 
\item ``null'' if the candidate set contains neither.
\end{enumerate}
In this article, we will always consider that a genomic location is in the
candidate set if and only if the read contains at least one seed with a
perfect match for this genomic location.
\end{definition}

With our assumptions, a read is mapped to the wrong location if and only
if the candidate set is off-target. Indeed, if the candidate set is null,
the read is not mapped, and if the candidate set is on target, the correct
location is discovered at the alignment step. The equivalence is granted
by the assumption that the true location is also the best, reducing the
mapping problem to a seeding problem. That being said, the true location
is not always the best in practical applications. We will show in
section~\ref{sec:truevsbest} that off-target seeding is responsible for
most of the mapping errors even without the assumption above, but for now
we maintain the strict equivalence between mapping errors and off-target
seeding.

Focusing on popular heuristics to map short reads, our aim is to develop a
method to estimate the probability that the candidate set is off-target.
Previous work pioneered a method to compute seeding probabilities but it
did not distinguish off-target from null
seeding~\cite{filion2017analytic,filion2018analytic}, and therefore did
not provide a way to estimate the frequency of mapping errors. Other
authors investigated the reliability of mapping
algorithms~\cite{pmid23872968}, but they focused on the probabilities of
random hits, recognizing that addressing the problem of incorrect mapping
requires taking into consideration the repeat structure of the genome.

The rest of the article is organized as follows: section~\ref{sec_seeds}
presents common seeding strategies used in bioinformatics,
section~\ref{sec:symbolic} presents the basic concepts of analytic
combinatorics that will be required to compute seeding probabilities,
sections~\ref{sec:offexact} to \ref{sec_offMEM} treat the cases of exact
seeds, skip seeds and MEM seeds, three common seeding heuristic used for
mapping, section~\ref{sec:sesame} presents Sesame, a C library
implementing the main results of the theory, section~\ref{sec_practical}
returns to the mapping problem and revisits the assumptions of the model,
and finally section~\ref{sec_disc} provides some perspectives on the
present work. Appendix~\ref{app_def} gathers for reference all the
definitions encountered in the text, and appendix~\ref{app_extra} contains
proofs and complements omitted from the main text.

@@

\begin{definition}
A seed is a subsequence of the read that has size at least $\gamma$
(defined by the context of the problem) and that has at least one perfect
match in the reference genome. Every genomic match of every seed is in the
candidate set. 
\end{definition}

@@

The success of the mainstream mappers BWA-MEM~\cite{li2013aligning} and
Bowtie2~\cite{pmid22388286} is due in part to the FM-index, which only
supports contiguous seeds. Some workarounds are
available~\cite{horton2008dislex, gagie2017compressed} but they increase
the memory footprint, explaining that short reads are typically mapped
using contiguous seeds. More generally, computing the sensitivity of
spaced seeds is challenging~\cite{pmid16819802, li2006superiority,
martin2017faster}. It is theoretically possible to do this using the
tools introduced below, but the complexity becomes prohibitive for seed
models with several don't-care positions. For these reasons we will not
treat the case of spaced seeds and we will give more detail in
section~\ref{sec_disc} about the kind of difficulties they represent for
our approach.

@@

Exact seeds are seeds of fixed size $\gamma$. In other words, when using
exact seeds, the candidate set consists of all the genomic locations for
which there is a perfect match of size $\gamma$ in the read. This seeding
heuristic was used in the first version of BLASTN~\cite{pmid2231712}, but
it has become unpopular for producing many short spurious hits.

Figure~\ref{fig:exact_seed_example} shows the exact seeds from an example
read with three miscalled nucleotides. The sequenced DNA fragment has
three duplicates so the seeds can match four possible locations.

@@

Before going further, we also need to move a practical consideration out
of the way. Seeds can match any sequence of the genome, not just the
target or the duplicates. However, we will ignore matches in the rest of
the genome because such random matches are unlikely to cause a mapping
error when seeding is off-target, contrary to matches in duplicates.
Neglecting those will greatly simplify the exposition of the theory
without loss of generality. We will explain in
section~\ref{sec:random_seeds} how to deal with this practical case and
how to identify those matches as off-target. Until then, we will consider
that the target and the duplicates are the only sequences in the reference
genome.

@@

To give an example, assume that a particular symbol, say $\Downarrow$, has
a probability of occurrence equal to $p$. The weighted generating function
words containing only this symbol is $pz$. The weight of the word is its
probability (here equal to $p$) and the size is its length (here 1).

In this document we will focus on the weighted generating function $A(z)$
of the set $\mathcal{A}$ of reads that do not have on-target seeds
(\textit{i.e.} reads for which seeding is either null or off-target). The
weight of a read is its probability of occurrence and the size $k$ is its
number of nucleotides. The coefficient $a_k$ is thus the proportion of
reads of size $k$ that do not have an on-target seed, which is the
quantity of interest.

@@

The analytic combinatorics framework relies on a strategy referred to as
the \emph{symbolic method}~\cite{sedgewick2013introduction}. The idea is
to combine simple objects into more complex objects. Each combinatorial
operation on the objects corresponds to a mathematical operation on their
weighted generating functions. One can thus obtain the weighted generating
function of complex objects, whose coefficients $a_k$ ($k \geq 0$) are the
probabilities of interest~\cite{regnier2000unified, nicodeme2002motif,
sedgewick2013introduction}.

@@

We define the mismatch alphabet $\mathcal{A}_0 = \{\square, |,
\Downarrow\}$, where $\square$ represents a correct nucleotide,
$\Downarrow$ represents an erroneous nucleotide and $|$ is a special
symbol appended after the last nucleotide to mark the end of the read. It
is not associated with a nucleotide and therefore has size 0.

@@

The probability of occurrence of the symbol $\Downarrow$ is $p$ (the error
rate of the sequencer) so the probability of occurrence of the $\square$
symbol is $1-p = q$. Both symbols have size 1, so their respective
weighted generating functions are $pz$ and $qz$. Using the rule for
concatenation, we see that a segment of $i$ symbols $\square$ followed by
a terminator has weighted generating function $(qz)^ipz$. The final symbol
$|$ has size 0, so a tail segment of $i$ symbols $\square$ followed by the
symbol $|$ has weighted generating function $(qz)^i$.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.85]{sketch_skip.pdf}
\caption{\textbf{The skip encoding.}
The read of Figure~\ref{fig:simple} is represented in the skip-3 alphabet.
The symbols $\Downarrow_j (j = 1,2,3)$ represent mismatches against the
target (they are erroneous nucleotides) and the $\square$ symbol
represents a match (it is a correct nucleotide). The vertical bars
indicate non-skipped positions (the potential start of a seed). The number
$j$ in $\Downarrow_j$ indicates the number of nucleotides until the next
non-skipped position. For $j=0$ the next nucleotide is not skipped.
Other features are as in Figure~\ref{fig:simple}.}
\label{fig:skip}
\end{figure}

@@

By construction, the Taylor expansion of the top right term in the matrix
$M_n(z)\cdot(I-M_n(z))^{-1}$ contains the probabilities of interest. More
specifically, if the Taylor series of this term is $a_0 + a_1 z + a_2 z^2
+ \ldots$, then $a_k$ is the probability that a read of size $k$ contains
no skip-$n$ seed of minimum size $\gamma$.

@@

We now turn our attention to the problem of computing the probability that
the seeding process is off-target when using exact seeds---recall from
section~\ref{sec:twotypes} that off-target seeding means that the
candidate set contains a duplicate but not the target.

If there is no duplicate (\textit{i.e.} $N = 0$), seeding cannot be
off-target, it can only be on target or null. So from here we assume that
the target has $N \geq 1$ duplicates. Let $S_0$ denote the event that
there is an on-target seed and let $S_j$ denote the event that there is a
seed for the $j$-th duplicate. We are thus interested in computing
$P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big)$, where
$\overline{S}_j$ denotes the complement of the event $S_j$. First observe
that
\begin{equation}
\label{eq:pcomp}
P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big) =
P(\overline{S}_0) - P\big(\overline{S}_0 \cap \overline{S}_1 \cap \ldots
\cap \overline{S}_N\big).
\end{equation}

Since the duplicates are assumed to evolve independently of each other and
through the same mutagenesis process, the events $S_j$ ($1 \leq j \leq N$)
are independent and identically distributed conditionally on
$\overline{S}_0$. We can thus write
\begin{eqnarray}
\notag
P(\overline{S}_0 \cap \ldots \cap \overline{S}_N) &=&
  P(\overline{S}_0) \cdot P(\overline{S}_1 \cap \ldots \cap
  \overline{S}_N | \overline{S}_0) \\
\label{eq:Pnull}
  &=& P(\overline{S}_0) \cdot P(\overline{S}_1 | \overline{S}_0)^N =
  P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0 \cap
\overline{S}_1)}{P(\overline{S}_0)} \right)^N.
\end{eqnarray}
Combining the two equations above, we obtain
\begin{equation}
\label{eq:Poff}
P\big(\overline{S}_0 \cap (S_1 \cup \ldots \cup S_N)\big) =
P(\overline{S}_0) - P(\overline{S}_0) \cdot \left( \frac{P(\overline{S}_0
\cap \overline{S}_1)}{P(\overline{S}_0)} \right)^N.
\end{equation}

Hence, the probability that seeding is off-target is a function of just
two quantities: $P(\overline{S}_0)$ and $P(\overline{S}_0 \cap
\overline{S}_1)$. The first is the probability that the read has no seed
for the target, which we have already computed in
section~\ref{sec:example_exact} using recursive expression
(\ref{eq:recur}). We now need to find a way to compute $P(\overline{S}_0
\cap \overline{S}_1)$.

\begin{remark}
Observe that expression (\ref{eq:Pnull}) is the probability of null
seeding (the read contains no seed for the target or any of its
duplicates). Since it is also a function of just $P(\overline{S}_0)$ and
$P(\overline{S}_0 \cap \overline{S}_1)$, it can be computed at no
additional cost. This probability is less relevant than the probabilities
that the seeding process is on target or off-target, but at times, it may
be useful to know the probability that a read is not mappable, especially
when reads are relatively short.
\end{remark}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.65]{curves_exact.pdf}
\caption{\textbf{Off-target seeding probabilities (exact seeds).} The
surfaces show the probability that seeding is off-target as a function of
the divergence rate $\mu$ and of the number $N$ of duplicates. The results
are shown for exact seeds of size $\gamma=19$ and for reads of size $k=50$
nucleotides sequenced with an error rate $p=0.01$. The divergence rate
$\mu$ is defined as the probability that a given duplicate differs from
the target at any given position. Note the factor 1000 in the scale of the
z-axis.}
\label{fig:curves_exact}
\end{figure}

Figure~\ref{fig:curves_exact} shows the result for a number of duplicates
$N$ from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The first
observation is that the probability that seeding is off-target increases
with $N$. This is also clear from expression (\ref{eq:Poff}). This can
also be understood intuitively because the probability of not seeding the
target is fixed and the probability of having an empty candidate set
decreases as $N$ increases. As a result the probability that the candidate
set contains only invalid candidates increases with $N$.

@@

To compute the probability that the seeding process is off-target when
using skip seeds, we observe that the logic of section~\ref{sec:offexact}
can be transposed with few modifications. In particular, the probability
can be computed through expression (\ref{eq:Poff}), where $S_0$ is the
event that the read has a \emph{skip} seed for the target (instead of an
exact seed) and $S_1$ is the event that the read has a \emph{skip} seed
for the first duplicate (instead of an exact seed).

@@

For the last two types of terminators, we must pay attention to the fact
that in general, a mismatch against $+$ that follows a mismatch against
$-$ can be represented by either a $\Downarrow_j$ symbol ($0 \leq j \leq
n$) or a $\downarrow_{/j}^+$ symbol ($j \geq 1$). The terminator is
$\Downarrow_j$ if the two mismatches are within the same interval between
non-skipped positions because both sequences locally have match length 0.
If the mismatch against $+$ is in another interval, the terminator is
$\downarrow_{/j}^+$ because in that case the $-$ sequence has a positive
match length. For now, bear in mind that a mismatch against the $+$
sequence only can produce a $\Downarrow_j$ terminator ($0 \leq j \leq n$),
this will be important later.

@@

We now have all the elements to specify the transfer matrix of reads with
no skip-$n$ seed for either sequence. Recall that $n$ is the number of
skipped nucleotides, $\gamma$ is the minimum seed length, $a$ is the
probability of a double match, $b$ is the probability of a mismatch
against the $-$ sequence only, $c$ is the probability of a mismatch
against the $+$ sequence only and $d$ is the probability of a double
mismatch. For notational convenience we define
\begin{eqnarray}
\label{eq_N}
N_i(z) &=& 1+z+\ldots+z^i, \\
W_j(z) &=& (az)^x \big( 1 + (az)^{n+1} + \ldots +
  (az)^{(n+1)m} \big) dz, \\
  \notag
  &\quad& \text{where } x = -j-1 \modulo{n+1},
  \text{ and } m = \left\lfloor
  \frac{\gamma-1-x}{n+1} \right\rfloor, \\
U_{i,j}(z) &=& (az)^x
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
\begin{cases}
  bz\cdot(az)^x \text{ if } j < y, \\
  0 \text{ otherwise}
\end{cases} \\
  \notag
  &\quad& \text{where } x = -i-j-1 \modulo{n+1},
  y = -i \modulo{n+1}, \\
  \notag
  &\quad& \text{and } m = \left\lfloor \frac{\gamma-1-i-x}{n+1}
    \right\rfloor, \\
V_{i,j}(z) &=& (az)^x 
  \big( 1 + (az)^{n+1} + \ldots + (az)^{(n+1)m} \big)dz +
\begin{cases}
  cz\cdot(az)^x \text{ if } j \leq x, \\
  0 \text{ otherwise}
\end{cases} \\
  \notag
  &\quad& \text{ where } x = -i-j-1 \modulo{n+1},
  m = \left\lfloor \frac{\gamma-1-i-x}{n+1} \right\rfloor.
\end{eqnarray}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.47]{curves_skip.pdf}
\caption{\textbf{Off-target seeding probabilities (skip seeds).}
The surfaces show the probability that seeding is off-target for skip-5
and skip-9 seeds of size $\gamma=19$. Read size and error rate are the
same as in Figure~\ref{fig:curves_exact}, \textit{i.e.} $k=50$ and
$p=0.01$. The divergence rate $\mu$, defined as the probability that a
given duplicate differs from the target at any given position. Note the
factor 1000 in the scale of the z-axis.}
\label{fig:curves_skip}
\end{figure}

Figure~\ref{fig:curves_skip} shows the result for a number of duplicates $N$
from 1 to 10 and for a divergence rate $\mu$ from 0 to 0.20. The surfaces
have the same general aspect as in Figure~\ref{fig:curves_exact}. The
probability that seeding is off-target increases with $N$ and there is
again a  worst value of $\mu$, because the maximum of $P(\overline{S}_0
\cap \overline{S}_1)$ minimizes expression (\ref{eq:Poff}) for every value
of $N$. However, those values of $\mu$ are not constant (they are
approximately 0.065 and 0.060 for skip-5 and skip-9 seeds, respectively).

Importantly, Figure~\ref{fig:curves_skip} reveals that skipping 5
nucleotides increases the chances that seeding is off-target by a factor
approximately 1.5, and skipping 9 nucleotides by a factor approximately
1.8 (compare with Figure~\ref{fig:curves_exact}). Skipping nucleotides
seems to have little effect, but this is not a general conclusion. Indeed,
skipping nucleotides decreases the probability of on-target seeding
(skipping positions implies fewer on-target seeds) but increases the
probability of null seeding (skipping positions implies fewer seeds
overall), so the effects must be evaluated on a case-by-case basis.
Skipping nucleotides can even decrease the probability that seeding is
off-target. Concretely, for exact seeds of size $\gamma=19$ in reads of
size $k=50$ with $N=10$ duplicates at divergence $\mu=0.1$ with a
sequencing error $p=0.1$, the probability of off-target seeding is
approximately 0.178 with exact seeds and 0.035 with skip-9 seeds, showing
skipping nucleotides can have different effects.
